---
title: "Practical 2: Compartmental Models to Equations"
author: "Sam Abbott"
output: 
      word_document:
      html_document:
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: resources/library.bib  
csl: resources/bmj.csl
---

```{r packages, include = FALSE}
library(tidyverse)
library(prettypublisher)
```

# Learning Objectives

1. How to translate a simple model flow diagram into equations
1. Understand how changing parameter values can change model dynamics
1. Using a simple model scaffold to develop a more complex model

# Timeline for Session

1. Set up (5 minutes)
1. Explore the dynamics of a simple SEIR model (10 minutes)
1. Add high and low risk latency to a SEIR model (10 minutes)
1. Explore model dynamics (10 minutes)
1. Extension: Translate your Model Flow Diagram to Equations (20 minutes)
1. Extension: Explore the parameter space of multiple models (20 minutes)(If having trouble with the previous excercises then skip to this point for an R free exercise)
1. Session wrap up (10 minutes)

# Excercises

## A Simple SEIR Model of TB

### Populations and Initialisation

```{r seir-initialisation}


```

### Parameters

```{r seir-parameters}

```

### Equations

```{r seir-equations}

```

### Simulate and Summarise

```{r seir-simulate}

```

```{r seir-summary-stats}

```

```{r seir-plot}


```

### Explore

- What impact does varying parameters have?

## Add High and Low Risk Compartments

- outline simple model function with text indicating where solution should go
- code to plot model etc

## Extension: Translate your Model Flow Diagram to Equations

- basic model outline
- suggestions for order in which to add complexity

## Extension: Explore the Parameter Space of Multiple Models

- link to shiny app
- help with shiny app
- question ideas

# References